---
title: "Habitat characteristics"
author: 
    - "Seba De Bona"
date: ""
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overwiew & Setup

This vignette explores the microhabitat types defined during the Density Manipulation Experiment and determines how different from each other they are based on the hydrological and ecological measures obtained from them.

```{r, warning=FALSE, message=FALSE}
# loading necessary packages
# for data handling
library(tidyverse)
library(magrittr)

# this requires the package guppyDme to be installed. If the user wishes not to install the package, please comment out the "library(guppyDme)" line and run instead the lines commented out below

#### If you have the guppyDme package installed
# loading package
library(guppyDme)

# #### If you do NOT wish to install the guppyDme package, please uncomment and run the following two lines of code, adding the package functions to the Global Environment and loading the data
# source(file.path(here::here(), "R", "package_functions.R"))
# load(file.path(here::here(), "data", "DMEhabitat.rda"))

# loading data
habitat <- DMEhabitat
```

## Habitat exploration

I will here produce some plot to compare microhabitat types and see if they are distninguishable.
I am attaching below a figure that depicts the 5 different habitat types, and the position/appearance in a pool.


```{r}
# scoring habitat to obtain numeric values. Using custom function substrate_score
habitat %<>%
  mutate(substrate_score = score_substrate(substrate))

# toggling with the parameter "conflicts"" in score_substrate() is possible to explore how
# only considering finer or coarser substrate type changes results.
# Please see the help documentation for the function.

# some of the habitat definitions are still a combination of many habitats (will be sorted later).
# for now, I will conly consider the 5 main habitat types when they appear alone

habitat %<>%
  filter(habitat %in% c("A", "B", "C", "D", "E")) %>%
  mutate(habitat = factor(habitat)) # removes empty factor levels

# defining color hues
# from https://projects.susielu.com/viz-palette?colors=[%22#ffd700%22,%22#ffb14e%22,%22#fa8775%22,%22#ea5f94%22,%22#cd34b5%22,%22#9d02d7%22,%22#0000ff%22]&backgroundColor=%22white%22&fontColor=%22black%22&mode=%22normal%22
color_hues <- c(
  "#ffd700",
  "#fa8775",
  "#cd34b5",
  "#9d02d7",
  "#0000ff")

# Plotting...

# Flow
p1 <- ggplot(habitat, aes(x=habitat, y=flow, fill = habitat)) + 
  geom_violin(scale = "area")
p1 + ggtitle("Flow across habitats") +
  xlab("Habitat") + ylab("Water speed (m/s)") +
  geom_boxplot(width=0.05, fill = "white") +
  scale_x_discrete(labels = hablab()) +
  scale_fill_manual(labels = hablab(), values = color_hues)

# Depth
p2 <- ggplot(habitat, aes(x=habitat, y=depth, fill = habitat)) + 
  geom_violin(scale = "area")
p2 + ggtitle("Depth across habitats") +
  xlab("Habitat") + ylab("Water depth (cm)") +
  geom_boxplot(width=0.05, fill = "white") +
  scale_x_discrete(labels = hablab()) +
  scale_fill_manual(labels = hablab(), values = color_hues)


# Substrate
p3 <- ggplot(habitat, aes(x=habitat, y=substrate_score, fill = habitat)) + 
  geom_violin(scale = "area")
p3 + ggtitle("Substrate coarseness") +
  xlab("Habitat") + ylab("Substrate type") +
  geom_boxplot(width=0.05, fill = "white") +
  scale_y_continuous(breaks = c(0,1,2,3,4), 
                     labels = c("silt", "sand", "gravel", "pebbles", "rock")) +
  scale_x_discrete(labels = hablab()) +
  scale_fill_manual(labels = hablab(), values = color_hues)

# Organic Matter
om <- habitat %>%
  group_by(habitat) %>%
  summarize(detritus = mean(detritus, na.rm = T),
            leaves = mean(leaves, na.rm = T))
om %<>%
  gather(key = organic_matter, value = p, detritus, leaves)

ggplot(om, aes(x = habitat, y = p, fill = organic_matter)) + 
  ggtitle("Organic matter presence") +
  xlab("Habitat") + ylab("Proportion present") +
  geom_bar(stat="identity", position=position_dodge()) +
  ylim(0, 1) +
  scale_fill_manual(values=c("#999999", "#339933")) +
  scale_x_discrete(labels = hablab())
```

It might be easier to visualize the average (and range of) values for each category in each habitat.

```{r}
habitat %>%
  select(streamID, habitat, depth, flow, substrate_score, detritus, leaves) %>%
  gather(key = variable, value = value, depth, flow, substrate_score, detritus, leaves) %>%
  group_by(habitat, variable) %>%
  summarize(mean_value = mean(value, na.rm = T),
            lo_q = quantile(value, 0.025, na.rm = T),
            hi_q = quantile(value, 0.975, na.rm = T)) %>%
  mutate(mean_value = round(mean_value, 2),
         lo_q = round(lo_q, 2),
         hi_q = round(hi_q, 2)) %>%
  add_column(bl = "[",
             br = "]") %>%
  unite(range, lo_q, hi_q, sep = " - ") %>%
  unite(range, bl, range, br, sep = "") %>%
  unite(value, mean_value, range, sep = " ") %>%
  spread(key = variable, value = value)
```

